---
title: "R Basics for Lecture 1 A"
font-family: 'Corbel'
output: github_document
---

## Econ B2000, Statistics and Introduction to Econometrics
## Kevin R Foster, Colin Powell School, the City College of New York, CUNY


```{r echo=FALSE}

load("acs2017_ny_data.RData")
attach(acs2017_ny)

```


## Variable Coding
Some of the PUMS variables here have a natural interpretation, for instance Age is measured in years. Actually even this has a bit of a twist, look at the histogram. 
```{r}
hist(AGE[(AGE > 90)])

```

There is a bit of weirdness in the right, where it looks like there are suddenly a bunch of people who are 95 but nobody is 94 or 96. This is due to a coding choice by the Census, where really old people are just labeled as "95" (top-coding) so it actually should be interpreted as meaning "92 or older". So if you were to get finicky (and every good statistician is!) you might go back to the calculations of averages previously and modify them all like this, to select just those who are female and who are coded as having age less than 90. Many variables are topcoded! *And recall that topcoding wouldn't change the median values calculated before, which is a point in favor of that statistic.*

```{r}
mean(AGE[ (female == 1) & (AGE<90) ]) 
```
You go make those other changes, figure out how top-coding changes the calculations of average age by gender -- I'll wait right here...

## Variable Coding Again
So we were saying that some variables, like Age - ahem! -- have a natural interpretation as a number.

Others are logical variables (called dummies) like female, Hispanic, or married - there is a yes/no answer that is coded 1/0. *Note that if you're creating these on your own it's good to give names that have that sort of yes/no answer, so a variable named 'female' is better than one named 'gender' where you'd have to remember who are coded as true and who are false.*

Many variables, like PUMA, have no natural explanation at all. Here are the first codes,
```{r}
str(as.numeric(PUMA))
```

You have to go to the codebook (or, in this case, the file PUMA_levels.csv or acs2017_codebook.txt from the zip file) to find out that 3801 codes for Washington Heights/Inwood, 3802 is Hamilton Heights/Manhattanville/West Harlem, etc. The program will happily calculate the average value for PUMA (type in *mean(PUMA)* and see for yourself!) but this is a meaningless value -- wtf is the average neighborhood code value!? If you want to select just people living in a particular neighborhood then you'd have to look at the list below. 

PUMA | Neighborhood
---- | ------------
3701 | NYC-Bronx CD 8--Riverdale, Fieldston & Kingsbridge
3702 | NYC-Bronx CD 12--Wakefield, Williamsbridge & Woodlawn
3703 | NYC-Bronx CD 10--Co-op City, Pelham Bay & Schuylerville
3704 | NYC-Bronx CD 11--Pelham Parkway, Morris Park & Laconia
3705 | NYC-Bronx CD 3 & 6--Belmont, Crotona Park East & East Tremont
3706 | NYC-Bronx CD 7--Bedford Park, Fordham North & Norwood
3707 | NYC-Bronx CD 5--Morris Heights, Fordham South & Mount Hope
3708 | NYC-Bronx CD 4--Concourse, Highbridge & Mount Eden
3709 | NYC-Bronx CD 9--Castle Hill, Clason Point & Parkchester
3710 | NYC-Bronx CD 1 & 2--Hunts Point, Longwood & Melrose
3801 | NYC-Manhattan CD 12--Washington Heights, Inwood & Marble Hill
3802 | NYC-Manhattan CD 9--Hamilton Heights, Manhattanville & West Harlem
3803 | NYC-Manhattan CD 10--Central Harlem
3804 | NYC-Manhattan CD 11--East Harlem
3805 | NYC-Manhattan CD 8--Upper East Side
3806 | NYC-Manhattan CD 7--Upper West Side & West Side
3807 | NYC-Manhattan CD 4 & 5--Chelsea, Clinton & Midtown Business District
3808 | NYC-Manhattan CD 6--Murray Hill, Gramercy & Stuyvesant Town
3809 | NYC-Manhattan CD 3--Chinatown & Lower East Side
3810 | NYC-Manhattan CD 1 & 2--Battery Park City, Greenwich Village & Soho
3901 | NYC-Staten Island CD 3--Tottenville, Great Kills & Annadale
3902 | NYC-Staten Island CD 2--New Springville & South Beach
3903 | NYC-Staten Island CD 1--Port Richmond, Stapleton & Mariner's Harbor
4001 | NYC-Brooklyn CD 1--Greenpoint & Williamsburg
4002 | NYC-Brooklyn CD 4—Bushwick
4003 | NYC-Brooklyn CD 3--Bedford-Stuyvesant
4004 | NYC-Brooklyn CD 2--Brooklyn Heights & Fort Greene
4005 | NYC-Brooklyn CD 6--Park Slope, Carroll Gardens & Red Hook
4006 | NYC-Brooklyn CD 8--Crown Heights North & Prospect Heights
4007 | NYC-Brooklyn CD 16--Brownsville & Ocean Hill
4008 | NYC-Brooklyn CD 5--East New York & Starrett City
4009 | NYC-Brooklyn CD 18--Canarsie & Flatlands
4010 | NYC-Brooklyn CD 17--East Flatbush, Farragut & Rugby
4011 | NYC-Brooklyn CD 9--Crown Heights South, Prospect Lefferts & Wingate
4012 | NYC-Brooklyn CD 7--Sunset Park & Windsor Terrace
4013 | NYC-Brooklyn CD 10--Bay Ridge & Dyker Heights
4014 | NYC-Brooklyn CD 12--Borough Park, Kensington & Ocean Parkway
4015 | NYC-Brooklyn CD 14--Flatbush & Midwood
4016 | NYC-Brooklyn CD 15--Sheepshead Bay, Gerritsen Beach & Homecrest
4017 | NYC-Brooklyn CD 11--Bensonhurst & Bath Beach
4018 | NYC-Brooklyn CD 13--Brighton Beach & Coney Island
4101 | NYC-Queens CD 1--Astoria & Long Island City
4102 | NYC-Queens CD 3--Jackson Heights & North Corona
4103 | NYC-Queens CD 7--Flushing, Murray Hill & Whitestone
4104 | NYC-Queens CD 11--Bayside, Douglaston & Little Neck
4105 | NYC-Queens CD 13--Queens Village, Cambria Heights & Rosedale
4106 | NYC-Queens CD 8--Briarwood, Fresh Meadows & Hillcrest
4107 | NYC-Queens CD 4--Elmhurst & South Corona
4108 | NYC-Queens CD 6--Forest Hills & Rego Park
4109 | NYC-Queens CD 2--Sunnyside & Woodside
4110 | NYC-Queens CD 5--Ridgewood, Glendale & Middle Village
4111 | NYC-Queens CD 9--Richmond Hill & Woodhaven
4112 | NYC-Queens CD 12--Jamaica, Hollis & St. Albans
4113 | NYC-Queens CD 10--Howard Beach & Ozone Park
4114 | NYC-Queens CD 14--Far Rockaway, Breezy Point & Broad Channel


Now you're probably thinking, isn't there some easier way? Yes there is. R has variables called "factors" that join together the long list of codes with a separate file telling what those codes mean. Later when we do further statistics, R will know how to appropriately treat these factors. (Also it will then give an error if you calculate mean(PUMA), which is proper.)
```{r}
PUMA <- as.factor(PUMA)
female <- as.factor(female)
```

I will leave you to worry over the recoding of the other variables, because it's good for the soul. I will show you 2 ways -- the quick and dirty way, and the fancy correct way.

First the quick and dirty way.
```{r}
print(levels(female))
levels(female) <- c("male","female")
```
Well, ways,
```{r}
educ_indx <- factor((educ_nohs + 2*educ_hs + 3*educ_somecoll + 4*educ_college + 5*educ_advdeg), levels=c(1,2,3,4,5),labels = c("No HS","HS","SmColl","Bach","Adv"))
```

(If you can figure out how that bit of code works, that would be good)

These just type in the levels. But for things like PUMA, it could be a long list and might not even match every one. To do it better, we need help from an R package.

###Detour on Packages
But first a bit of a detour, to mention how to use packages. R depends crucially on "packages" - that's the whole reason that the open-source works. Some statistician invents a cool new technique, then writes up the code in R and makes it available. If you used a commercial program you'd have to wait a decade for them to update it; in R it's here now. Also if somebody hacks a nicer or easier way to do stuff, they write it up.

So enter this into the Console,
```
install.packages("tidyverse")
install.packages("plyr")
```
then
```{r message=FALSE}
library(tidyverse)
library(plyr)
levels_n <- read.csv("PUMA_levels.csv")
levels_orig <- levels(PUMA) 
levels_new <- join(data.frame(levels_orig),data.frame(levels_n))
levels(PUMA) <- levels_new$New_Level
```

Alt, from R-Studio, click "Tools" then "Install Packages..." and tell it to install the packages, "plyr" and "tidyverse". That is nice if you want to see some of the packages or if you don't quite remember the name. Then the next piece of code, library, tells the program that you want to use commands from this package.

Those commands read in a little csv file that I had made, with the PUMA codes, then matches the old codes with the new complete text. Note that I'm lazy so codes in NY state outside of NYC are coded NA.

###Back from Detour
R will do the summary differently when it knows the variable is a factor,
```{r}
summary(female)
summary(PUMA)
summary(educ_indx)
```
To find mean and standard deviation by neighborhood, you could use something like this,

```{r}
ddply(acs2017_ny, .(PUMA), summarize, mean = round(mean(AGE), 2), sd = round(sd(AGE), 2), n_obsv = length(PUMA))
```
Although tapply would also work fine.

Here's the 90th and 10th percentiles of wages by neighborhood,

```{r}
dat_use1 <- subset(acs2017_ny,((INCWAGE > 0) & in_NYC))
ddply(dat_use1, .(PUMA), summarize, inc90 = quantile(INCWAGE,probs = 0.9), inc10 = quantile(INCWAGE,probs = 0.1), n_obs = length(INCWAGE))
```

You could also use table (or crosstabs) for factors with fewer items,

```{r}
table(educ_indx,female)
xtabs(~educ_indx + female)
```
Want proportions instead of counts?
```{r}
prop.table(table(educ_indx,female))
```
*Remember prop.table later when we do marginals.*

Try it and see what happens if you use table with PUMA...

This data includes not just whether a person has a college degree but also what field was the degree in: Economics or Psychology, for instance. Look over the codebook about DEGFIELD and DEGFIELDD (that second D means more detail) to see the codes. Maybe look at 10th and 90th percentiles by degree field?

In general, R is very flexible so there are often many different ways to get the same answer. There are some people who love to debate which is best. (Often, tradeoff between speed and intelligibility.) For now just worry about learning at least one way. Later on you can go back and refine your techniques.

Sometimes attaching a dataset makes things easier. But as you get more advanced you might find it better to include the dataset name inside the function. There are advantages and disadvantages each way and some of the intro texts suggest one or the other.

If you do a lot of analysis on a particular subgroup, it might be worthwhile to create a subset of that group, so that you don't have to always add on logical conditions. These two sets of expressions, looking at "prime-age" people, get the same results:

```{r message=FALSE}
mean(educ_nohs[(AGE >= 25)&(AGE <= 55)])
mean(educ_hs[(AGE >= 25)&(AGE <= 55)])
mean(educ_somecoll[(AGE >= 25)&(AGE <= 55)])
mean(educ_college[(AGE >= 25)&(AGE <= 55)])
mean(educ_advdeg[(AGE >= 25)&(AGE <= 55)])

# alternatively
restrict1 <- as.logical((AGE >= 25)&(AGE <= 55))
dat_age_primeage <- subset(acs2017_ny, restrict1)

detach()
attach(dat_age_primeage)

mean(educ_nohs)
mean(educ_hs)
mean(educ_somecoll)
mean(educ_college)
mean(educ_advdeg)

detach()


```
So you detach the original data frame and instead attach the restricted version. Then any subsequent analysis would be just done on that subset. Just remember that you've done this (again, this is a good reason to save the commands in a program so you can look back) otherwise you'll wonder why you suddenly don't have any kids in the sample! 

## Why All These Details?
You might be tired and bored by these details, but note that there are actually important choices to be made here, even in simply defining variables. Take the fraught American category of "race". This data has a variable, RACED, showing how people chose to classify themselves, as 'White,' 'Black,' 'American Indian,' 'Asian,' various combinations, and many more codes.

Suppose you wanted to find out how many Asians are in a particular population. You could count how many people identify themselves as Asian only; you could count how many people identify as Asian in any combination. Sometimes the choice is irrelevant; sometimes it can skew the final results (e.g. the question in some areas, are there more African-Americans or more Hispanics?).

Again, there's no "right" way to do it because there's no science in this peculiar-but-popular concept of "race". People's conceptions of themselves are fuzzy and complicated; these measures are approximations.

## Basics of government race/ethnicity classification
The US government asks questions about people's race and ethnicity. These categories are social constructs, which is a fancy way of pointing out that they are based on people's own views of themselves (influenced by how we think that other people think of us...). Currently the standard classification asks people separately about their "race" and "ethnicity" where people can pick labels from each category in any combination.

The "race" categories include: "White," "African-American," "American Indian," "Asian," and others. Then the supplemental race categories offer more detail.

These are a peculiar combination of very general (well over 40% of the world's population is "Asian") and very specific ("American Indian" has fewer) representing a peculiar history of popular attitudes in the US. Only in the 2000 Census did they start to classify people in mixed races. If you were to go back to historical US Censuses from more than a century ago, you would find that the category "race" included separate entries for Irish and French. Stephen J Gould has a fascinating book, The Mismeasure of Man, discussing how early scientific classifications of humans tried to "prove" which nationalities/races/groups were the smartest. Ta-Nehisi Coates notes, "racism invented race in America." Throughout history, statistics have been used to try to prove peoples' prejudices.

Note that "Hispanic" is not "race" but rather ethnicity (includes various other labels such as Spanish, Latino, etc.). So a respondent could choose "Hispanic" and any race category -- some choose "White," some choose "African American," some might be combined with any other of those complicated racial categories.

If you wanted to create a variable for those who report themselves as African-American and Hispanic, you'd use the expression (AfAm == 1) & (Hispanic == 1); sometimes government stats report for non-Hispanic whites so (white == 1) & (Hispanic != 1). You can create your own classifications depending on what questions you're investigating. This data includes items on birthplace and ancestry (more detail on relatives!).

The Census Bureau gives more information [here](http://www.census.gov/newsroom/minority_links/minority_links.html). 

All of these racial categories make some people uneasy: is the government encouraging racism by recognizing these classifications? Some other governments choose not to collect race data. But that doesn't mean that there are no differences, only that the government doesn't choose to measure any of these differences. In the US, government agencies such as the Census and BLS don't generally collect data on religion.

### Re-Coding complicated variables from initial data
If we want more combinations of variables then we create those. Usually a statistical analysis spends a lot of time doing this sort of housekeeping - dull but necessary. It has a variety of names: data carpentry, data munging...

Educational attainment is also classified with complicated codes in this data: the original data has code 63 to mean high school diploma, 64 for a GED, 65 for less than a year of college, etc. I have transformed them into a series of dummy variables, zero/one variables for whether a person has no high school diploma, just a high school diploma, some college, a bachelor's degree, or an advanced degree. As with race, there is no rule that you must always do it thus. Your own analysis might be different. (Is it valid to lump together everybody who lacks a high school diploma, no matter whether they completed just first grade or up to 12th?)

That's the whole point of learning to do the data work for yourself: you can see all of the little decisions that go into creating a conclusion. Some conclusions might be fragile so a tiny decision about coding could change everything; other conclusions are robust to deviations. You must find out.

# De-bugging
Without a doubt, programming is tough. In R or with any other program, it is frustrating and complicated and difficult to do it the first few times. Some days it feels like a continuous battle just to do the simplest thing! Keep going despite that, keep working on it. 

Your study group will be very helpful of course.

Often a google search of the error message helps. If you've isolated the error and read the help documentation on that command, then you're on your way to solving the problem on your own.

If you have troubles that you can't solve, email me for help. But try to narrow down your question: if you run 20 lines of code that produce an error, is there a way to reproduce the error in just 5 lines? What if you did the same command on much simpler data, would it still cause an error? Sending emails like "I have a problem with errors" might be cathartic but is not actually useful to anyone. If you email me with the minimal code that recreates the error, along with the text of the error and/or a screenshot, then that will help more.

##Do it
The first homework assignment asks you to start working on these questions. Begin by running the code that I give here, just to see if you can replicate my results. Then start asking more questions about the data - there is so much info there! Immigration/ancestry status, how they commute, health insurance, poverty, income, owner or renter (and how much they pay!), all sorts of info. Have some fun.

```{r echo=FALSE}
detach()
```
